---
title: "Untitled"
author: "Margaret Janiczek"
date: "2025-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## R Markdown

```{r}
dat <- read.csv("inpatient.csv", sep = "|")

dat_outpatient <- read.csv("outpatient.csv", sep = "|")

dat_nursing <- read.csv("snf.csv", sep = "|")
# download icd10


firearm_codes <- c('W32', 'W33', 'X72', 'X73', 'X74',
                       'X93', 'X94', 'X95',  
                       'Y22', 'Y23', 'Y24', 
                       'Y35', 'Z04')

firearm_names <- list(
  "W32" = "Accidental handgun discharge",
  "W33" = "Accidental discharge of larger firearm (rifle, shotgun)",
  "X72" = "Intentional self-harm by handgun discharge",
  "X73" = "Intentional self-harm by larger firearm discharge",
  "X74" = "Intentional self-harm by other firearm",
  "X93" = "Assault by handgun discharge",
  "X94" = "Assault by larger firearm discharge (rifle, shotgun)",
  "X95" = "Assault by unspecified firearm discharge",
  "Y22" = "Handgun discharge, undetermined intent",
  "Y23" = "Rifle/shotgun/large firearm discharge, undetermined intent",
  "Y24" = "Unspecified firearm discharge, undetermined intent",
  "Y35" = "Injuries resulting from legal intervention",
  "Z04" = "Examination encounter"
)

firearm_key <- data.frame(
  ICD_SHORT = names(firearm_names),
  icd_descrip = unlist(firearm_names), 
  category = c("Accidental", "Accidental", "Self-harm", "Self-harm",
               "Self-harm", "Assault", "Assault", "Assault", "Undetermined", "Undetermined", "Undetermined", "Legal intervention", "Other")
)

dat_long <- pivot_longer(dat,
                         cols = starts_with("ICD_DGNS"), 
                         values_to = "ICD_CODE")

dat_long_snf <- pivot_longer(dat_nursing,
                         cols = starts_with("ICD_DGNS"), 
                         values_to = "ICD_CODE") %>%
  mutate(ICD_SHORT = str_sub(ICD_CODE, start = 1L, end = 3L))


dat_long_sub <- dat_long %>%
  mutate(ICD_SHORT = str_sub(ICD_CODE, start = 1L, end = 3L))

dat_long_outpatient <- dat_outpatient %>%  select(BENE_ID, CLM_ID, starts_with("ICD_DGNS")) %>%
  pivot_longer(          cols = starts_with("ICD_DGNS"), 
                         values_to = "ICD_CODE")



dat_long_sub_outpatient <- dat_long_outpatient %>%
  mutate(ICD_SHORT = str_sub(ICD_CODE, start = 1L, end = 3L))

dat_long_sub_nursing <- dat_long_snf %>%
  mutate(ICD_SHORT = str_sub(ICD_CODE, start = 1L, end = 3L))%>%
  filter(ICD_SHORT %in% firearm_codes)

dat_long_sub_outpatient <- dat_long_sub_outpatient %>%
  filter(ICD_SHORT %in% firearm_codes)



dat_long_sub_inpatient <- dat_long_sub %>% 
  filter(ICD_SHORT %in% firearm_codes)

length(unique(dat_long_sub_inpatient$BENE_ID))


firearm_injury_sub_in <- unique(dat_long_sub_inpatient$BENE_ID)
firearm_injury_sub_out <- unique(dat_long_sub_outpatient$BENE_ID)

firearm_injury_all <- unique(c(firearm_injury_sub_in, firearm_injury_sub_out))

length(unique(dat$CLM_ID))
```

```{r}
# estimating costs of claims



dat_analysis_in <- dat %>%
  filter(CLM_ID %in% dat_long_sub_inpatient$CLM_ID) %>%
  group_by(CLM_ID) %>%
  slice(1) %>%
  left_join(dat_long_sub_inpatient %>% dplyr::select(BENE_ID, CLM_ID, name, ICD_CODE, ICD_SHORT) %>% distinct()) %>%
  left_join(firearm_key)

dat_analysis_out <- dat_outpatient %>%
  filter(CLM_ID %in% dat_long_sub_outpatient$CLM_ID) %>%
  group_by(CLM_ID) %>%
  slice(1)%>%
  left_join(dat_long_sub_outpatient %>% dplyr::select(BENE_ID, CLM_ID, name, ICD_CODE, ICD_SHORT) %>% distinct()) %>%
  left_join(firearm_key)

length(unique(dat_analysis_in$BENE_ID))

length(unique(dat_analysis_in$CLM_ID))

length(unique(dat_long_sub_outpatient$BENE_ID))

length(unique(dat_long_sub_outpatient$CLM_ID))

# calculate total cost
```

```{r}
# summarize initial data

library(gtsummary)
library(tidyverse)

dat_analysis_in %>%
  ungroup() %>%
  dplyr::select(
    CLM_TOT_CHRG_AMT, CLM_PMT_AMT,
    icd_descrip
  ) %>%
  tbl_summary()


dat_analysis_out %>%
  ungroup() %>%
  dplyr::select(
    CLM_TOT_CHRG_AMT, CLM_PMT_AMT,
    icd_descrip
  ) %>%
  tbl_summary()
```



```{r}
date_of_firearms <- dat_analysis_in %>%
  ungroup() %>%
  dplyr::select(BENE_ID, firearm_clm_date = CLM_FROM_DT) %>%
  bind_rows(dat_analysis_out %>% ungroup() %>% dplyr::select(BENE_ID, firearm_clm_date = CLM_FROM_DT)) %>%
  mutate(firearm_clm_date = dmy(firearm_clm_date))


# patients before and after

dat_sub_inpatients <- dat %>%
  filter(BENE_ID %in% c(dat_analysis_in$BENE_ID, dat_analysis_out$BENE_ID)) %>%
  group_by(CLM_ID) %>%
  slice(1) %>%
  select(BENE_ID, CLM_ID, CLM_TOT_CHRG_AMT, CLM_PMT_AMT, CLM_FROM_DT, everything()) %>%
  mutate(
    firearm_event = CLM_ID %in% dat_analysis_in$CLM_ID
  ) %>%
  left_join(date_of_firearms)

dat_sub_inpatients <- dat_sub_inpatients %>%
  mutate(
    before_or_after = dmy(CLM_FROM_DT) < firearm_clm_date
  )

# now get whether events were before or after firearm event



dat_sub_outpatients <- dat_outpatient %>%
  filter(BENE_ID %in% c(dat_analysis_out$BENE_ID, dat_analysis_in$BENE_ID)) %>%
  group_by(CLM_ID) %>%
  slice(1)

dat_sub_nursing <- dat_nursing %>%
  filter(BENE_ID %in% c(dat_analysis_out$BENE_ID, dat_analysis_in$BENE_ID)) %>%
  group_by(CLM_ID) %>%
  slice(1)


length(unique(dat_sub_inpatients$BENE_ID))
```

Principal diagnosis code for many of the inpatient claims was "Awaiting organ transplant".


???What is Claim Line Number? 

?? Calculating total cost of visit

Calc cost of claim or of line item? 


# Introduction


# Methods


# Results

There were 58066 inpatient claims. Of these, 112 (0.2%) inpatient claims had an ICD-10 code related to firearms injury, which came from 23 unique beneficiary IDs. 

There were additionally 575,092 outpatient claims, of which 30 (<0.001%) had an ICD-10 code related to firearms injury, from 13 unique beneficiaries. 

