---
title: "Untitled"
author: "Margaret Janiczek"
date: "2025-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## R Markdown

```{r}
dat <- read.csv("inpatient.csv", sep = "|")

dat_outpatient <- read.csv("outpatient.csv", sep = "|")

dat_nursing <- read.csv("snf.csv", sep = "|")
# download icd10


firearm_codes <- c('W32', 'W33', 'X72', 'X73', 'X74',
                       'X93', 'X94', 'X95',  
                       'Y22', 'Y23', 'Y24', 
                       'Y35', 'Z04')

firearm_names <- list(
  "W32" = "Accidental handgun discharge",
  "W33" = "Accidental discharge of larger firearm (rifle, shotgun)",
  "X72" = "Intentional self-harm by handgun discharge",
  "X73" = "Intentional self-harm by larger firearm discharge",
  "X74" = "Intentional self-harm by other firearm",
  "X93" = "Assault by handgun discharge",
  "X94" = "Assault by larger firearm discharge (rifle, shotgun)",
  "X95" = "Assault by unspecified firearm discharge",
  "Y22" = "Handgun discharge, undetermined intent",
  "Y23" = "Rifle/shotgun/large firearm discharge, undetermined intent",
  "Y24" = "Unspecified firearm discharge, undetermined intent",
  "Y35" = "Injuries resulting from legal intervention",
  "Z04" = "Examination encounter"
)

firearm_key <- data.frame(
  ICD_SHORT = names(firearm_names),
  icd_descrip = unlist(firearm_names), 
  category = c("Accidental", "Accidental", "Self-harm", "Self-harm",
               "Self-harm", "Assault", "Assault", "Assault", "Undetermined", "Undetermined", "Undetermined", "Legal intervention", "Other")
)

dat_long <- pivot_longer(dat,
                         cols = starts_with("ICD_DGNS"), 
                         values_to = "ICD_CODE")

dat_long_snf <- pivot_longer(dat_nursing,
                         cols = starts_with("ICD_DGNS"), 
                         values_to = "ICD_CODE") %>%
  mutate(ICD_SHORT = str_sub(ICD_CODE, start = 1L, end = 3L))


dat_long_sub <- dat_long %>%
  mutate(ICD_SHORT = str_sub(ICD_CODE, start = 1L, end = 3L))

dat_long_outpatient <- dat_outpatient %>%  select(BENE_ID, CLM_ID, starts_with("ICD_DGNS")) %>%
  pivot_longer(          cols = starts_with("ICD_DGNS"), 
                         values_to = "ICD_CODE")



dat_long_sub_outpatient <- dat_long_outpatient %>%
  mutate(ICD_SHORT = str_sub(ICD_CODE, start = 1L, end = 3L))

dat_long_sub_nursing <- dat_long_snf %>%
  mutate(ICD_SHORT = str_sub(ICD_CODE, start = 1L, end = 3L))%>%
  filter(ICD_SHORT %in% firearm_codes)

dat_long_sub_outpatient <- dat_long_sub_outpatient %>%
  filter(ICD_SHORT %in% firearm_codes)



dat_long_sub_inpatient <- dat_long_sub %>% 
  filter(ICD_SHORT %in% firearm_codes)

length(unique(dat_long_sub_inpatient$BENE_ID))


firearm_injury_sub_in <- unique(dat_long_sub_inpatient$BENE_ID)
firearm_injury_sub_out <- unique(dat_long_sub_outpatient$BENE_ID)

firearm_injury_all <- unique(c(firearm_injury_sub_in, firearm_injury_sub_out))

length(unique(dat$CLM_ID))
```

```{r}
# estimating costs of claims



dat_analysis_in <- dat %>%
  filter(CLM_ID %in% dat_long_sub_inpatient$CLM_ID) %>%
  group_by(CLM_ID) %>%
  slice(1) %>%
  left_join(dat_long_sub_inpatient %>% dplyr::select(BENE_ID, CLM_ID, name, ICD_CODE, ICD_SHORT) %>% distinct()) %>%
  left_join(firearm_key)

dat_analysis_out <- dat_outpatient %>%
  filter(CLM_ID %in% dat_long_sub_outpatient$CLM_ID) %>%
  group_by(CLM_ID) %>%
  slice(1)%>%
  left_join(dat_long_sub_outpatient %>% dplyr::select(BENE_ID, CLM_ID, name, ICD_CODE, ICD_SHORT) %>% distinct()) %>%
  left_join(firearm_key)

length(unique(dat_analysis_in$BENE_ID))

length(unique(dat_analysis_in$CLM_ID))

length(unique(dat_long_sub_outpatient$BENE_ID))

length(unique(dat_long_sub_outpatient$CLM_ID))

# calculate total cost
```

```{r}
# summarize initial data

library(gtsummary)
library(tidyverse)

dat_analysis_in %>%
  ungroup() %>%
  dplyr::select(
    CLM_TOT_CHRG_AMT, CLM_PMT_AMT,
    icd_descrip
  ) %>%
  tbl_summary()


dat_analysis_out %>%
  ungroup() %>%
  dplyr::select(
    CLM_TOT_CHRG_AMT, CLM_PMT_AMT,
    icd_descrip
  ) %>%
  tbl_summary()
```



Captured date of firearms incident. 

```{r}

# date of FIRST firearm visit
date_of_firearms <- dat_analysis_in %>%
  ungroup() %>%
  dplyr::select(BENE_ID, firearm_clm_date = CLM_FROM_DT) %>%
  # grab dates from outpatient too
  bind_rows(dat_analysis_out %>% ungroup() %>% dplyr::select(BENE_ID, firearm_clm_date = CLM_FROM_DT)) %>%
  mutate(firearm_clm_date = dmy(firearm_clm_date)) %>%
  group_by(BENE_ID) %>%
  arrange(firearm_clm_date) %>%
  slice(1)  %>%
  ungroup()


# patients before and after

dat_sub_inpatients <- dat %>%
  filter(BENE_ID %in% c(dat_analysis_in$BENE_ID, dat_analysis_out$BENE_ID)) %>%
  group_by(CLM_ID) %>%
  slice(1) %>%
  select(BENE_ID, CLM_ID, CLM_TOT_CHRG_AMT, CLM_PMT_AMT, CLM_FROM_DT, everything()) %>%
  mutate(
    firearm_event = CLM_ID %in% dat_analysis_in$CLM_ID
  ) %>%
  left_join(date_of_firearms)

dat_sub_inpatients <- dat_sub_inpatients %>%
  mutate(
    before_or_after = dmy(CLM_FROM_DT) < firearm_clm_date
  )

# now get whether events were before or after firearm event

# TODO get other out and nursing claims as above 

dat_sub_outpatients <- dat_outpatient %>%
  filter(BENE_ID %in% c(dat_analysis_out$BENE_ID, dat_analysis_in$BENE_ID)) %>%
  group_by(CLM_ID) %>%
  slice(1) %>%
  select(BENE_ID, CLM_ID, CLM_TOT_CHRG_AMT, CLM_PMT_AMT, CLM_FROM_DT, everything()) %>%
  mutate(
    firearm_event = CLM_ID %in% dat_analysis_in$CLM_ID
  ) %>%
  left_join(date_of_firearms)%>%
  mutate(
    before_or_after = dmy(CLM_FROM_DT) < firearm_clm_date
  )

dat_sub_nursing <- dat_nursing %>%
  filter(BENE_ID %in% c(dat_analysis_out$BENE_ID, dat_analysis_in$BENE_ID)) %>%
  group_by(CLM_ID) %>%
  slice(1)%>%
  select(BENE_ID, CLM_ID, CLM_TOT_CHRG_AMT, CLM_PMT_AMT, CLM_FROM_DT, everything()) %>%
  mutate(
    firearm_event = CLM_ID %in% dat_analysis_in$CLM_ID
  ) %>%
  left_join(date_of_firearms)%>%
  mutate(
    before_or_after = dmy(CLM_FROM_DT) < firearm_clm_date
  )


length(unique(dat_sub_inpatients$BENE_ID))
```

# Calculate cost of claims before and after incident

```{r}
dat_sub_inpatients %>%
  filter(firearm_event == FALSE) %>%
  ungroup() %>%
  dplyr::select(
    CLM_TOT_CHRG_AMT, before_or_after
  ) %>%
  tbl_summary(by = before_or_after)

dat_sub_outpatients %>%
  filter(firearm_event == FALSE) %>%
  ungroup() %>%
  dplyr::select(
    CLM_TOT_CHRG_AMT, before_or_after
  ) %>%
  tbl_summary(by = before_or_after)

dat_sub_nursing %>%
  filter(firearm_event == FALSE) %>%
  ungroup() %>%
  dplyr::select(
    CLM_TOT_CHRG_AMT, before_or_after
  ) %>%
  tbl_summary(by = before_or_after)
```


Principal diagnosis code for many of the inpatient claims was "Awaiting organ transplant".


???What is Claim Line Number? 

?? Calculating total cost of visit

Calc cost of claim or of line item? 

# Abstract 
In the United States, between 2009 and 2017, there was an average of more than 85,000 emergency department visits for firearms-related injuries, and in 2019 firearms accounted for more deaths on average than deaths due to motor-vehicle accidents. Not only is the toll on human life significant, but also there are financial ramifications for both individuals and the healthcare system. An emergency-room visit related to firearms injury has been estimated to cost above \$5,000 on average (\$70,000 if the injury was severe enough to warrant hospital admission), and patient costs in the 6 months after firearm injury are on average found to be 3 times the cost compared to the 6 months prior to firearm injury. In this study we will aim to replicate these findings in the available CMS synthetic dataset. First, estimate the average cost of claim for a firearms-related injury, then estimate the average change in claims comparing the year before to the year after injury. We will compare these estimates based on grouping states by various regulatory buckets: minimum age to purchase (no minimum, 18, 21), concealed carry laws (unrestricted, permit, banned), and looking at states with and without assault weapon bans. We will also set up the framework for answering causal inference questions regarding whether implementation of additional gun laws decreases the amount and cost of firearm-related injuries. We will use states which implemented new gun laws in the wake of the 2018 Parkland shooting which include Florida, Vermont, New Jersey, Washington, Pennsylvania, and Oregon.


# Introduction


# Methods


# Results

There were 58066 inpatient claims. Of these, 112 (0.2%) inpatient claims had an ICD-10 code related to firearms injury, which came from 23 unique beneficiary IDs. 

There were additionally 575,092 outpatient claims, of which 30 (<0.001%) had an ICD-10 code related to firearms injury, from 13 unique beneficiaries. 

